---
title: 'Exploratory Data Analysis:'
subtitle: 'Presidential Campaign Contribution in Ohio '
---
Haleh Dolati
=============================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Loading all of the packages that I'm going to use in my analysis 
library(data.table)
library(gapminder)
library(ggplot2)
library(GGally)
library(plotly)
library(reshape2)
library(RColorBrewer)
library(Hmisc)
library(dplyr)
library(gridExtra)
library(grid)
library(ggplot2)
library(lattice)
library(ggplot2)
library(tidyr)
library(rgeos)
library(gpclib)
library(maptools)
library(sp)
library(rgdal)
library(scales)
library(ggmap)
library(zoo)
library(mapproj)
```

```{r echo=FALSE, warning=FALSE}
# Load the Data
#setwd('P4-submission')
election<-read.csv("OH_Election.csv")

```

###Introduction
> I chose the “Financial Contributions to Presidential Campaigns” data for the
State of Ohio. I chose Ohio because I went to The Ohio Stated University and
am fairly familiar with the state's political dynamics. Ohio is considered a 
swing state, which makes it very interesting to study. This data set has the 
data for 167,259 donations.

###Summary of Data Set 
> Now that I know what kind of data I have in this dataset, I would like to
get some general information, like data types.


```{r echo=FALSE, warning=FALSE}
str(election)
```
###Initial Observation from the Summary
> This dataset contains individuals’ financial contributions to 2016
presidential candidates.  It shows whom, how much, when, and where contributed
to every single candidate during both primary and general elections.  
However, there are some variables in the contribution data that do not give 
us any insight like the candidates ID number. So I decided to keep 
only 9 variables: "cmte_id","cand_id","contbr_st","receipt_desc","memo_cd",
"memo_text","form_tp","file_num", and "tran_id"

```{r echo=FALSE, warning=FALSE}
election <- election[,!names(election) %in%
                    c("cmte_id","cand_id","contbr_st","receipt_desc",
                      "memo_cd","memo_text","form_tp","file_num","tran_id")]
str(election)
```

> I also noticed the zip codes probably need some cleaning. So I'm going to 
first make sure only 5 digit zip codes are present in the variable. I will 
replace the unfit zip codes with NA. 

```{r echo=FALSE, warning=FALSE}
election$contbr_zip <- substring(election$contbr_zip, 1, 5)
summary(as.numeric(election$contbr_zip))
```

> Since the Min and max zip codes are out of range, I've decided to convert 
any zip code that does not start with 43, 44, and 45 since all zip codes in 
Ohio should start with these numbers. The first zip code is 43001 and the 
last one is 45999.  

```{r echo=FALSE, warning=FALSE}
in_ohio <- as.character(seq(43000, 46000,1))
election$contbr_zip <- ifelse(election$contbr_zip %in% in_ohio, 
                              election$contbr_zip, 'NA')
summary(as.numeric(election$contbr_zip))
```

> Now that I cleaned up the zip code variable, the first questions that came 
to my mind were: who received more money overall? Which party received more 
money? How contributions changed general and primary elections?  However, 
I’ve noticed the party affiliation is missing fro this dataset.  Therefore, 
if I want to answer couple of these questions I need to add that to the 
dataset. First I need a list of candidates: 

```{r echo=FALSE, warning=FALSE}
cat("Candidates: ", "\n")
levels(election$cand_nm)
```

> Based on my initial though on visualization and analysis, I need to add two
other variables. The first one is a new variable for each candidate's party
affiliation.  
The second one is a categorical one (based on the donation amount) for the 
level of contribution so I can see if there is any differences between number 
of small and big donors regarding the candidate that they chose. I chose the
following categories:   
[0-$50) [$50-$100) [$100, $500) [$500, $2700) [$2700, )  
My next step is to apply this grouping to all three datasets: election, 
primary and general.  

```{r echo=FALSE, warning=FALSE, Load_the_Data}
Democrat <- c("Clinton, Hillary Rodham", 
              "Sanders, Bernard", 
              "O'Malley, Martin Joseph", 
              "Lessig, Lawrence", 
              "Webb, James Henry Jr.")
Republican <- c("Bush, Jeb", "Carson, Benjamin S.",
                 "Christie, Christopher J.", "Cruz, Rafael Edward 'Ted'",
                 "Fiorina, Carly", "Graham, Lindsey O.",
                 "Huckabee, Mike", "Jindal, Bobby",
                 "Kasich, John R.", "Pataki, George E.",
                 "Paul, Rand", "Perry, James R. (Rick)",
                 "Rubio, Marco", "Santorum, Richard J.",
                 "Trump, Donald J.", "Walker, Scott")
Libertarian <- c("Johnson, Gary")
Independent <- c("McMullin, Evan" )
GreenParty <- c("Stein, Jill")

election$party <- ifelse(election$cand_nm %in% Democrat, "Democrat",
        ifelse(election$cand_nm %in% Republican, "Republican",
        ifelse(election$cand_nm %in% Libertarian, "Libertarian",
        ifelse(election$cand_nm %in% GreenParty, "GreenParty",
        ifelse(election$cand_nm %in% Independent, "Independent",
                        NA  )))))
```

```{r echo=FALSE, warning=FALSE}
group_donation <- function(contb_receipt_amt) {
                  Contributions <- as.integer(contb_receipt_amt)
                  if (Contributions < 50) {
                    return("[0-$50)")
                  } else if (Contributions < 100) {
                    return("[$50-$100)")
                  } else if (Contributions < 500) {
                    return("[$100, $500)")
                  } else if (Contributions <= 2700) {
                    return ("[$500, $2700)")
                  } else {
                    return ("[$2700, )")
                  }
              }


# apply the function to create a new column in election
election$donation_level <- apply(election, 1, function(x) 
  group_donation(x["contb_receipt_amt"]))
# convert that column to ordered factors, so we can sort
election$donation_level <- factor(election$donation_level, 
                             levels = c("[0-$50)", "[$50-$100)", 
                                        "[$100, $500)", "[$500, $2700)", 
                                        "[$2700, )"))
```
> This data set has information on when each donation were made. In order to 
make it ready for plotting, I have to convert the contb_receipt_dt (date of 
donation) into a date type variable so I can track contributions over time. 
The original format is 11-Jul-16 and I'm going to convert them to 2016-07-11.

```{r echo=FALSE, warning=FALSE}
election$contb_receipt_dt <- as.Date(election$contb_receipt_dt, "%d-%b-%y")

```

> And finally I've noticed was the data has the information on contributions 
for both primary and general elections. I would like to see how donation has 
changed from primary to general election so the next step for me is to make 
two separate data frames for them: primary_election and general_election.

```{r echo=FALSE, warning=FALSE}
primary_election <- subset(election, election_tp=='P2016')
general_election <- subset(election, election_tp=='G2016')
```
# Univariate Plots Section

> My first question is: How much Ohioan contributed to the 2016 presidential 
election? But before I answer this question I took a look at the contribution 
receipt amount data and found out there are some negative contributions 
(probably refunds). I manually checked couple of those negative records and 
found out there are positive records from the same contributor for the exact 
same amount but with a positive sign. Both records for these cancelled 
transactions need to be excluded. Here is the new summary of election data:

```{r echo=FALSE, warning=FALSE}
election$abs_amount <- abs(election$contb_receipt_amt)
setorder(election, 'abs_amount', 'contbr_nm', 'contb_receipt_dt', 
         'contb_receipt_amt')
election_sum <- rollapply(zoo(election$contb_receipt_amt), 2, sum)
election_sum <- c(1, election_sum)
election$sum <- election_sum
election$sum <- ifelse(election$contb_receipt_amt < 0, 0, election$sum)

election <- subset(election, sum != 0 )
str(election)
```

```{r echo=FALSE, warning=FALSE}
total_donation = sum(election$contb_receipt_amt)
primary_donation = sum(primary_election$contb_receipt_amt)
general_donation = sum(general_election$contb_receipt_amt)
cat("Total dollars donated during 2016 presidential election:", 
    sum(election$contb_receipt_amt), "\n")
cat("Total dollars donated 2016 presidential primary election:", 
    sum(primary_election$contb_receipt_amt), "\n")
cat("Total dollars donated 2016 presidential general election:", 
    sum(general_election$contb_receipt_amt), "\n")
cat("Total count of donations:", length(election$contb_receipt_amt), "\n")
cat("Total number of donors:", length(unique(election$contbr_nm)))
```

> Well, majority of the contributions happened during the primaries. I also 
noticed the sum of contributions during primary and general election is lower 
than total contributions. I checked the data and it is because there were 
other types of denotable election type in the state of Ohio during the 
presidential race. Let’s take a look at the distribution of these donations 
based on the amount of them.

```{r echo=FALSE, warning=FALSE}
total_donation_dist <- ggplot(aes(x = contb_receipt_amt), data = election) + 
  geom_histogram(binwidth = 500)+ 
  scale_y_log10()+
  labs(title='Total')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

primary_donation_dist <- ggplot(aes(x = contb_receipt_amt), 
                                data = primary_election) +
  scale_y_log10()+
  geom_histogram(binwidth = 500)+
  labs(title='Primary')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

general_donation_dist <- ggplot(aes(x = contb_receipt_amt), 
                                data = general_election) + 
  scale_y_log10()+
  geom_histogram(binwidth = 500)+ 
  labs(title='General')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(total_donation_dist, primary_donation_dist, 
             general_donation_dist, ncol=3)
```

> It looks like majority of the donations were small donations. However, maximum legal contribution by individuals is $2700. I'm going to take a look at those donations that were made by individuals, not the corporations. Also I'm not interested in refunds, so the negative ones are going to be excluded as well.

```{r echo=FALSE, warning=FALSE}
total_donation_dist_2700 <- ggplot(aes(x = contb_receipt_amt), data = election) +
  geom_histogram(binwidth = 100)+
  labs(title='Total')+
  scale_x_continuous(limits = c(0, 2700),breaks = seq(0, 2700, 500)) +
  scale_y_continuous(limits = c(0, 30000),breaks = seq(0, 30000, 5000))
primary_donation_dist_2700 <- ggplot(aes(x = contb_receipt_amt), 
                                     data = primary_election) + 
  geom_histogram(binwidth = 100)+
  labs(title='Primary')+
  scale_x_continuous(limits = c(0, 2700),breaks = seq(0, 2700, 500)) +
  scale_y_continuous(limits = c(0, 20000),breaks = seq(0, 20000, 5000))
general_donation_dist_2700 <- ggplot(aes(x = contb_receipt_amt), 
                                     data = general_election) + 
  geom_histogram(binwidth = 100)+
  labs(title='General')+
  scale_x_continuous(limits = c(0, 2700),breaks = seq(0, 2700, 500)) +
  scale_y_continuous(limits = c(0, 15000),breaks = seq(0, 15000, 5000))
total_donation_dist_2700
primary_donation_dist_2700
general_donation_dist_2700

```

  > So overall, the majority of donations were under $100. What about the donations that each party and each candidate received from individual Ohioan (under $2700)?
(Range of number of donations for different parties are significantly different. Therefore, I’m going to use logarithm (log10) to scale the numbers. )

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = contb_receipt_amt), data = election) + 
  geom_histogram(binwidth = 50)+ 
  scale_x_continuous(limits = c(0, 2700),breaks = seq(0, 2700, 1000)) +
  scale_y_log10()+
  facet_wrap(~party)
```

```{r echo=FALSE, warning=FALSE}
ggplot(data = election, aes( x = cand_nm)) +
  geom_bar() +  
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```



```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = contb_receipt_amt), data = election) + 
  geom_histogram(binwidth = 50)+ 
  scale_x_continuous(limits = c(0, 2700),breaks = seq(0, 2700, 1000)) +
  scale_y_continuous(trans='log10')+
  facet_wrap(~cand_nm)

```

> All I can say from these of plots is:  
1- Overall, Democrats and republicans received the majority of the money  
2- Clinton, Sanders, Trump, Cruz, Carson, and Kasich are the top 5 candidates 
in terms of donations to their campaigns.  
The last variable that I would like to take a look at is the date. I would 
like to see how the number of contributions changed during the election year:


```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=contb_receipt_dt), data=election) +
    geom_histogram(binwidth = 7) 
```

> After primaries more donations were make. The next chart I would like to see
is how number of donations were changed for each candidate? 

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=contb_receipt_dt), data=election) +
    geom_histogram(binwidth = 7)+
  scale_y_log10()+
  facet_wrap(~cand_nm)
```

> Naturally trump and Clinton received more donations for the general election
(after primaries) but Sanders also received many donations before the general 
election.  

# Univariate Analysis  
### What is the structure of your dataset?  
> Federal Election Commission collects data on every single donation and refund
to all candidates for all elections. The data that I'm using provides detailed
information on 167259 donations that were made during 2016 presidential 
election (primary and general) to 24 candidates. There are 18 variables for 
each contribution. To understand the data better, I also downloaded the 
“Data Description” that the Federal Election Commission's website provided. 
The followings are the description of each column:

Abbreviation  | Description | Type
------------- | ------------- | -------------
CMTE_ID | COMMITTEE ID | S  
CAND_ID | CANDIDATE ID | S  
CAND_NM | CANDIDATE NAME | S  
CONTBR_NM | CONTRIBUTOR NAME | S  
CONTBR_CITY | CONTRIBUTOR CITY | S  
CONTBR_ST | CONTRIBUTOR STATE | S  
CONTBR_ZIP | CONTRIBUTOR ZIP CODE | S  
CONTBR_EMPLOYER | CONTRIBUTOR EMPLOYER | S  
CONTBR_OCCUPATION | CONTRIBUTOR OCCUPATION | S  
CONTB_RECEIPT_AMT | CONTRIBUTION RECEIPT AMOUNT | N  
CONTB_RECEIPT_DT | CONTRIBUTION RECEIPT DATE | D	  
RECEIPT_DESC | RECEIPT DESCRIPTION | S
MEMO_CD | MEMO CODE | S
MEMO_TEXT | MEMO TEXT	 | S
FORM_TP | FORM TYPE | S
FILE_NUM | FILE NUMBER | N
TRAN_ID | TRANSACTION ID | S

### What is/are the main feature(s) of interest in your dataset?
> The main features of interest are the number of contributions, recipients of 
those contributions, the amount of the contributions, recipients of those
contributions, and the contributors. 

### What other features in the dataset do you think will help support your 
###investigation into your feature(s) of interest?
> The occupation of the contributors, the cities and the zip codes that the 
contributions are coming from.

### Did you create any new variables from existing variables in the dataset?
> I created two new variables: party and donation_level. Party is the 
candidates’ affiliated political party. Donation level is a categorical 
variable that shows the level of donation in 5 categories:  
[0-$50), [$50-$100), [$100, $500), [$500, $2700), [$2700, )

### Of the features you investigated, were there any unusual distributions? 
###Did you perform any operations on the data to tidy, adjust, or change the 
###form of the data? If so, why did you do this?
> I change the date's data type from 11-Jul-16 to 2016-07-11. This way I can 
plot the contribution over time. Also I cleaned up the zip code variable to 5 
digit format. 

# Bivariate Plots Section
> I would like to see the relationship between contribution amount and party 
and candidates. So I'm going to take a look at the box plot:  

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = party, y = contb_receipt_amt, fill = party), data = election) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 2700)) +
  scale_fill_manual(values = c('blue','green',  'gold', 'pink','red'))
```

> Although I just looked and small donations (under $2700) still the variety 
of donation amount are too much. I'm going to take a look at the log of them 
first then I will use other plot types to explore them further.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = party, y = contb_receipt_amt, fill = party), data = election) +
  geom_boxplot() +
  scale_y_log10() +
  scale_fill_manual(values = c('blue','green',  'gold', 'pink','red'))
```

> It looks the median contribution amount is lower for democrats compare to 
others and there are so many outliers for both republicans and democrats.  
My next question is How much money each party received overall, during primary 
and general?

```{r echo=FALSE, warning=FALSE}

group_by_party<- election %>%
  group_by(party) %>%
  summarise(mean_cand_nm=mean(contb_receipt_amt),
            median_cand_nm=median(contb_receipt_amt),
            sum_contribution = sum(contb_receipt_amt),
            Percent_contb= sum(contb_receipt_amt)*100/total_donation,
            n=n())%>%
  ungroup()
group_by_party_primary<- primary_election %>%
  group_by(party) %>%
  summarise(mean_cand_nm_p = mean(contb_receipt_amt),
            median_cand_nm_p = median(contb_receipt_amt),
            sum_contribution_p = sum(contb_receipt_amt),
            Percent_contb_p = sum(contb_receipt_amt)*100/total_donation,
            n=n())%>%
  ungroup()
#
group_by_party_general<- general_election %>%
  group_by(party) %>%
  summarise(mean_cand_nm_g = mean(contb_receipt_amt),
            median_cand_nm_g = median(contb_receipt_amt),
            sum_contribution_g = sum(contb_receipt_amt),
            Percent_contb_g = sum(contb_receipt_amt)*100/total_donation,
            n=n())%>%
  ungroup()
group_by_party_P_G <- merge(group_by_party_primary, group_by_party_general, 
                            by="party")
group_by_party_P_G_A <- merge(group_by_party, group_by_party_P_G, by="party")
group_by_party_P_G_A
```

```{r echo=FALSE, warning=FALSE}
slices <- group_by_party_P_G_A$sum_contribution
lbls <- group_by_party_P_G_A$party
pie(slices, labels = lbls, main="Pie Chart of Campaign Contributions", 
    col = rainbow(length(slices)))
legend("topright", c("Republican", "GreenParty","Democrat",
                     "Libertarian" ), cex = 0.7,
   fill = rainbow(length(slices)))

```

> The pie chart clearly shows that the two major political parties received the majority 
of the monetary contributions. However, since the amount of contribution is extreamly different between the political parties, the other 3 values are almost hidden. Inorder to see them better, I'm going to make a bar chart with log10 of values:

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x=party, y=sum_contribution), data=group_by_party_P_G_A)+
  geom_histogram(stat= "identity")

```

> The bar charts highlights the difference even more.  Since the other parties did not receive that 
much money, I'm going to compare only Republicans and Democrats for primary 
and general election.


```{r echo=FALSE, warning=FALSE}
sub_for_bar <- data.frame(group_by_party_P_G_A$party,
                          group_by_party_P_G_A$sum_contribution_p,
                          group_by_party_P_G_A$sum_contribution_g)
names(sub_for_bar) <- c("party", "primary", "general")

sub_for_bar <- subset(sub_for_bar, 
                    party=='Democrat' | party=='Republican')
melted <- melt(sub_for_bar, id.var= "party")
ggplot(melted, aes(x=party, y=value, fill= variable))+
  geom_bar(stat="identity")
```





> Now I want to know how different is the amount of money that each candidate 
received from Ohioans over all? What about the average donation amount?

```{r echo=FALSE, warning=FALSE}
group_by_candidate<- election %>%
  group_by(cand_nm) %>%
  summarise(mean_cand_nm=mean(contb_receipt_amt),
            median_cand_nm=median(contb_receipt_amt),
            sum_contribution = sum(contb_receipt_amt),
            n=n())

sum_contribution <- ggplot(data = group_by_candidate, 
                           aes(x = reorder(cand_nm, -sum_contribution), 
                               y = sum_contribution)) +
  geom_bar(stat = "identity") +
  labs(y = "Total Donations", 
       title = "Candidate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mean_contribution <- ggplot(data = group_by_candidate, 
                            aes(x = reorder(cand_nm, -mean_cand_nm), 
                                y = mean_cand_nm)) +
  geom_bar(stat = "identity") +
  labs(y = "Total Donations", 
       title = "Candidate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mean_contribution
sum_contribution
```

> It looks like three candidates received way more money than the others. 
Overall Ohioan donated more money to Hillary Clinton's campaign than any other
candidate. Apparently Kasich and Bush supporters donated more money on average
per donation but O'Malley and Pataki are not far behind. George Pataki received
the least amount of money from them, which is really interesting since Pataki's
supporters donated on average more money than many other candidates. In 
contrast, Sanders has the forth place when it comes to the total contribution 
but he has the lowest mean contribution!  
Another question is who from each party received the most money? To answer this
question, we only need to find the candidate from Republican Party who received
he most money: Clinton received the most money overall, therefore she got the 
most money among democrats as well.  Libertarian, Independent, and Green Party
had only one candidate so there is nothing to compare. So now I'm going to find
among republicans, who received the most monetary contribution?


```{r echo=FALSE, warning=FALSE}
Rebuplican_party_cand <- subset(election, party=='Republican')
Rebuplican_group <- Rebuplican_party_cand %>%
  group_by(cand_nm) %>%
  summarise(sum_contribution = sum(contb_receipt_amt),
            n=n())
Rebuplican_group$cand_nm[which.max( Rebuplican_group$sum_contribution )]

```

> Kasich received the most amounts of contributions, which makes sense since he
is the governor of Ohio and Pataki received the least amount of money among 
republicans.
But another thing came to my attention: Kasich received all those money during 
the primary election, while contribution to Trump and Clinton's campaign 
continued since they were nominees. So I would like to see primary and general 
election contribution stats separately: 
 


```{r echo=FALSE, warning=FALSE}
primary_group_by_candidate<- primary_election %>%
  group_by(cand_nm) %>%
  summarise(primary_mean_cand_nm=mean(contb_receipt_amt),
            primary_median_cand_nm=median(contb_receipt_amt),
            primary_sum_contribution = sum(contb_receipt_amt),
            n=n())
general_group_by_candidate<- general_election %>%
  group_by(cand_nm) %>%
  summarise(general_mean_cand_nm=mean(contb_receipt_amt),
            general_median_cand_nm=median(contb_receipt_amt),
            general_sum_contribution = sum(contb_receipt_amt),
            n=n())

primary <- ggplot(data= primary_group_by_candidate, 
                  aes(x=cand_nm, y=primary_sum_contribution))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(labels = scales::comma)

general <- ggplot(data= general_group_by_candidate, 
                  aes(x=cand_nm, y=general_sum_contribution))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(labels = scales::comma)

all <- ggplot(data= group_by_candidate, 
              aes(x=cand_nm, y=sum_contribution))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_y_continuous(labels = scales::comma)
all
primary
general
```

> I don’t think these plots going to help me! The scales are different on 
Y-axis and some candidates did not receive any contribution after primaries. 
What if I put genral and primary election together in one plot? 



```{r echo=FALSE, warning=FALSE}
cand_pg_compare <- merge( general_group_by_candidate, primary_group_by_candidate , by='cand_nm', all = TRUE)


cand_pg_compare <- subset( cand_pg_compare, select = -c(primary_mean_cand_nm,primary_median_cand_nm,general_mean_cand_nm, general_median_cand_nm, n.x, n.y) )

gp_melted <- melt(cand_pg_compare, id.var= "cand_nm")
ggplot(gp_melted, aes(x=cand_nm, y=value, color= variable ))+
  geom_point(stat="identity")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="top")
```



> Now I can see and compare them better. Interesting point is even after 
primaries, people continued to donate to some republican or democrat candidates
beside trump and Clinton. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(aes(x=contb_receipt_dt, y=contb_receipt_amt), data=election) +
  geom_point(alpha = 0.01) +
  geom_smooth(formula = contb_receipt_amt ~ contb_receipt_dt)+
  scale_y_continuous(trans = "log10")
```

> The refunds and corporation donations make it difficult to detect a pattern 
here. So I'm going to take look at individual donations.

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(aes(x=contb_receipt_dt, y=contb_receipt_amt), 
       data=subset(election, contb_receipt_amt <2700)) +
  geom_point(alpha=0.01, position='jitter') +
  geom_smooth(formula = contb_receipt_amt ~ contb_receipt_dt)+
  scale_y_continuous(trans = "log10")
```

> it looks with time individual contributors donated smaller and smaller amount
of money. Also it is interesting to see how donating to different candidates 
has changed during the lection. Specifically, I would like to see how it 
changed for the top 4 donation recipients.

```{r echo=FALSE, warning=FALSE}

candidate_groups_date <- group_by(election, cand_nm, contb_receipt_dt)
election_candidate_date <- summarise(candidate_groups_date, 
                                        sum_contbr = sum(contb_receipt_amt),
                                        mean_contbr = mean(contb_receipt_amt),
                                        median_contbr = median(contb_receipt_amt)
                                        )

ggplot(data = subset(election_candidate_date, 
                     cand_nm =="Trump, Donald J."|
                     cand_nm =="Clinton, Hillary Rodham" | 
                       cand_nm =="Sanders, Bernard" | 
                       cand_nm =="Kasich, John R."  
                         ), 
       aes(contb_receipt_dt, sum_contbr)) +
  geom_line(aes(color = cand_nm), alpha= 0.8) +
  scale_x_date() +
  scale_y_continuous( trans = "log10") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="top")+
  labs(y = "Total Donations") 
```

> Another variable that I would like to investigate is the contributor's 
occupation for individual donors. It looks for individual donors retired and 
attorneys  donated the highest amount of money! I'm a little surprised by 
almost half a million donation by not employed individuals! 

```{r echo=FALSE, warning=FALSE}
ind_contrb <- subset(election, contb_receipt_amt <2700)
group_by_occupation<- ind_contrb %>%
  group_by(contbr_occupation) %>%
  summarise(occup_sum_contribution = sum(contb_receipt_amt),
            n=n())
head(arrange(group_by_occupation,desc(occup_sum_contribution)), n = 5)
```
> And finally I would like to see how contributions has changes accross Ohio 
based on zipcodes:

```{r echo=FALSE, warning=FALSE}
group_by_zip <-  election %>%
  group_by(contbr_zip) %>%
  summarise(zip_sum_contribution = sum(contb_receipt_amt),
            mean_sum_contribution = mean(contb_receipt_amt),
            n=n())
group_by_zip = na.omit(group_by_zip)
head(arrange(group_by_zip,desc(zip_sum_contribution)), n = 5)
head(arrange(group_by_zip,desc(n)), n = 5)
head(arrange(group_by_zip,desc(mean_sum_contribution)), n = 5)
```


> There are a couple of interesting facts about donations based on zip codes:  
1- zip code 44122 (Cleveland area) the highest amount of total donations, which
is interesting because neither average donation nor number of donations in this
zip code are not the highest in the State.  
2- The highest amount of individual contribution happened in two zip code with 
only 1 donations. I looked up those two: 43659 is very small zip code in Toledo 
and 45262 is a P.O.Box. The next highest mean belongs to another small zip code
in east of Ohio with only 4 donations. 
There are 73 zip codes with only 1 donations. While I did not investigate all 
73 of them, the hands full that I looked up were either one-block zip codes or 
P.O.Boxes. The highest number of donations belongs to zip code 43214 in 
Columbus with 2068 donations. I would like to see the distribution of donation 
based on sum of donations and number of donations.


```{r echo=FALSE, warning=FALSE}
zip <- readOGR(dsn = "Ohio_zip_map", layer = "zip_code_area")
zip <- fortify(zip, region="ZIP_CODE")
group_by_zip$contbr_zip <- as.numeric(group_by_zip$contbr_zip)
```

```{r echo=FALSE, , warning=FALSE}
ggplot() +
  geom_map(data = group_by_zip, aes(map_id = contbr_zip, 
                                    fill = zip_sum_contribution), map = zip) +
  expand_limits(x = zip$long, y = zip$lat)+
  scale_y_continuous(breaks=NULL)+
  scale_x_continuous(breaks=NULL)+
  theme(axis.title.x=element_blank())+
  theme(axis.title.y=element_blank())+
  scale_fill_gradient(low = "lightyellow", high = "darkgreen")
```

> This maps show three significantly darker green blue areas: Columbus, 
Cincinnati, and Cleveland, which is not surprising since these three are 
largest city in Ohio. But I don’t get much more than this from it. So I'm going
to compare donations to Democrats and Republicans across Ohio.

```{r echo=FALSE, , warning=FALSE}
election_dems <- subset(election, party=='Democrat')
group_by_zip_dems <-  election_dems %>%
  group_by(contbr_zip) %>%
  summarise(zip_sum_contribution = sum(contb_receipt_amt),
            mean_sum_contribution = mean(contb_receipt_amt),
            n=n())

election_rep <- subset(election, party=='Republican')
group_by_zip_rep <-  election_rep %>%
  group_by(contbr_zip) %>%
  summarise(zip_sum_contribution = sum(contb_receipt_amt),
            mean_sum_contribution = mean(contb_receipt_amt),
            n=n())

```


```{r echo=FALSE, , warning=FALSE}
D <- ggplot() +
  geom_map(data = group_by_zip_dems, aes(map_id = contbr_zip, fill = n), 
           map = zip) +
  expand_limits(x = zip$long, y = zip$lat)+
  scale_y_continuous(breaks=NULL)+
  scale_x_continuous(breaks=NULL)+
  theme(axis.title.x=element_blank())+
  theme(axis.title.y=element_blank())+
    scale_fill_gradient(low = "lightblue", high = "darkblue",
                        breaks = c(200, 400, 600, 800, 1000, 
                                   1200,1400, 1600, 1800, 2000))+
  coord_fixed()
D
```

```{r echo=FALSE, , warning=FALSE}
R <- ggplot() +
  geom_map(data = group_by_zip_rep, aes(map_id = contbr_zip, fill = n), 
           map = zip) + expand_limits(x = zip$long, y = zip$lat)+
  scale_y_continuous(breaks=NULL)+ scale_x_continuous(breaks=NULL)+
  theme(axis.title.x=element_blank())+ theme(axis.title.y=element_blank())+
  scale_fill_gradient(low = "lightpink", high = "red",
                      breaks = c(200, 400, 600, 800, 1000, 
                                 1200,1400, 1600, 1800, 2000))+ coord_fixed()

R
```

> Bothe parties received more money from high density urban areas republican 
also received high number donation in lower-density zip codes as well. 

# Bivariate Analysis
### Talk about some of the relationships you observed in this part of the 
###investigation. How did the feature(s) of interest vary with other features 
###in the dataset?

> 58% of the total contribution in Ohio went to the Republican Party. Since the
number of Republican Party's candidates were 3 times more than democrats (5 
Democrats and 16 Republicans) this difference makes sense. However, in each 
party few candidates received the majority of the contributions. 
Another observation is for the primary election, democrats received more 
donations than the republicans, but for general elections republicans received 
majority of the donations.   
Finally, It looks like retired donated the majority of the money during the 
2016 presidential election.

### Did you observe any interesting relationships between the other features?
> I found it really interesting that the unemployed folks donated half a 
million to the campaigns! 

# Multivariate Plots Section

> Based what I learned about the donations, I wonder how the size of donation 
is different for each candidate. 
 

```{r echo=FALSE, , warning=FALSE}
top_candidate <- c("Clinton, Hillary Rodham", "Sanders, Bernard", 
                   "Carson, Benjamin S.", "Kasich, John R.", 
                   "Trump, Donald J.", "Cruz, Rafael Edward 'Ted'")

election_top_6 <- subset(election, cand_nm %in% top_candidate)

ggplot(data = election_top_6, aes(contb_receipt_amt)) +
  geom_histogram(binwidth = 50, aes(fill = cand_nm)) +
  scale_x_continuous(limits=c(0, 2700), breaks = seq(0, 2750, 250)) +
  labs(x = "Donation Amount", title = "Donations under $2700") +
  scale_y_continuous(breaks = seq(0, 50000, 5000), limits=c(0,50000))


```
 
> The most common donation amount is $50 for all candidates. I'm going to make 
the charts more specific. I would like to see these contributions for top 6 as 
well as Clinton and Trump, for donations under $100.

```{r echo=FALSE, warning=FALSE}
two_cand <- subset(election, (cand_nm == "Clinton, Hillary Rodham" | 
                                cand_nm =="Trump, Donald J.") &
                     (election_tp == "G2016" | election_tp == "P2016") ,
                   na.rm = TRUE) 

p1 <- ggplot(data = election_top_6,
             aes(contb_receipt_amt)) +
  geom_histogram(binwidth = 5, aes(fill = cand_nm)) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(x = "Donation Amount", title = "Donations under $100") +
  scale_y_continuous(breaks = seq(0, 30000, 5000))

p2 <- ggplot(data= two_cand,
             aes(contb_receipt_amt)) +
  geom_histogram(binwidth = 5, aes(fill = cand_nm)) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(x = "Donation Amount", title = "Donations under $100") +
  scale_y_continuous(breaks = seq(0, 30000, 5000))
p1
p2
``` 

> Ok! It looks $25 donation is the most common donation. Clinton Also received 
the smallest donations as well. Did the donation amount's pattern changed 
during the campaign year for these two candidates? 

```{r echo=FALSE, warning=FALSE}
ggplot(data=two_cand, aes(x = (contb_receipt_dt))) +
   geom_point(aes(y = (two_cand$contb_receipt_amt), colour = two_cand$cand_nm),
              show.legend = TRUE,  size=0.5, alpha=0.5)
```

```{r echo=FALSE, warning=FALSE}
ggplot(data = two_cand,
             aes(x=cand_nm, y= contb_receipt_amt, fill = election_tp)) +
  geom_bar(stat="identity") 

```
 
 > The time series plot shows how trump's donations increased when he got the 
 Republican Parties nomination. The stacked bar makes it really clear that 
 while Trump received significantly less money during the primaries than 
 Clinton, for the general election he was able to gather more donations. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the 
###investigation. Were there features that strengthened each other in terms of 
###looking at your feature(s) of interest? Were there any interesting or surprising
###interactions between features?  
> Almost all of the relationships where predictable: more donations will be made
closer to primary and general election day.
Two major parties will receive more money than others as well as two nominees 
from those parties. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE, Plot_One}
ggplot(data = subset(election_candidate_date, 
                     cand_nm =="Clinton, Hillary Rodham" | 
                       cand_nm =="Sanders, Bernard" | 
                       cand_nm =="Carson, Benjamin S." | 
                       cand_nm =="Kasich, John R." | 
                       cand_nm =="Trump, Donald J." | 
                       cand_nm =="Cruz, Rafael Edward 'Ted'" ), 
       aes(contb_receipt_dt, sum_contbr)) +
  geom_line(aes(color = cand_nm)) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b-%y")+ 
  scale_y_continuous(trans = "log10") +
  labs(y = "Total Donations in $", x= "Date", 
       title = "Candidate Donations Over Time, Top 6 Candidates")+
  theme(legend.position="bottom", legend.title = element_blank(),
        axis.text.x = element_text(angle = 40, hjust = 1))

```

### Description One
> This plot shows the monetary contribution over time to the top 6 candidates 
in terms of over all donation. Sanders's campaign received money even after 
2016 Democratic National Convention and nomination of Clinton as Democrats 
candidate. It might be because his supporters were hoping to convince him to 
run as an independent candidate. 
This plot also shows interesting spikes in Trump and Kasich’s campaign in some 
points. For Trump is it easy to explain: most of the donations happened after 
he became Republican's nominee. For Kasich, the most noticeable spike happened 
on mid December. 


### Plot Two
```{r echo=FALSE, warning=FALSE, Plot_Two}
ggplot(data = election_top_6,
             aes(contb_receipt_amt)) +
  geom_histogram(binwidth = 5, aes(fill = cand_nm)) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 5)) +
  labs(x = "Donation Amount", title ="Donations under $100, Top 6 Candidates ",
       y ="Number of Donations") +
  scale_y_continuous(breaks = seq(0, 30000, 5000))+
  theme(legend.position="bottom", legend.title = element_blank(),
        axis.text.x = element_text(angle = 40, hjust = 1))
```

### Description Two
> This plot shows the under $100 donation for the top 6 donation recipients 
candidates. I found this plot very interesting since it shows clearly what 
sanders said during his campaign: average donation to his campaign was $27. 
I think majority of his supporter donated somewhere around $25 so the mode of 
donations to his campaign got so close to the average. 

### Plot Three
```{r echo=FALSE, Plot_Three}
D <- ggplot() +
  geom_map(data = group_by_zip_dems, aes(map_id = contbr_zip, fill = n), 
           map = zip) + expand_limits(x = zip$long, y = zip$lat)+
  scale_y_continuous(breaks=NULL) + scale_x_continuous(breaks=NULL) +
  theme(axis.title.x = element_blank(), axis.title.y=element_blank(),
        legend.position='bottom' )+labs(title = "Democrats") + 
  scale_fill_gradient(low = "lightblue", high = "darkblue")+
  coord_fixed()

R <- ggplot() +
  geom_map(data = group_by_zip_rep, aes(map_id = contbr_zip, fill = n), 
           map = zip) + expand_limits(x = zip$long, y = zip$lat)+
  scale_y_continuous(breaks=NULL)+scale_x_continuous(breaks=NULL)+
  theme(axis.title.x=element_blank(), axis.title.y=element_blank(),
        legend.position='bottom')+
  labs(title = "Republican")+
    scale_fill_gradient(low = "lightpink", high = "red")+
  coord_fixed()
grid.arrange(R, D, top = "Number of Donations to Political Parties", 
             ncol = 2)
```

### Description Three  

> I like this visualization because it shows how different areas donated 
differently to Democrats and Republicans. While the darkness of the colors may 
represent almost the same amount of contribution for both parties in some areas,
but it is important to pay attention to the legend: the scales are extremely 
different. I tried to force equal breaks for these two maps but unfortunately 
it did not work. 

# Reflection

> Ohio 2016 presidential campaign contributions data has information on more 
than 160000 campaign donations July 2014 to December 2016. While I enjoyed 
working with such a large dataset and on a topic that I really interested in, 
since this was my first big project in R faced some challenges.

###Challenge and My Solutions  

> •	The first big problem that I faced was a simple data cleaning for the zip 
codes. After a lot of search I found substring() will solve part of my problem 
in just 1 line of code.  
•	I wanted to add a party variable to my dataset so I can summarize the data and
visualize it based on political parties. I spend a good amount of time googling
and learning about ifelse() and %in% which helped me to add that variable.  
•	But the challenging part of data cleaning was removing the refunded 
contributions. I looked into many Stack Overflow post, watched many YouTube 
tutorials, looked into similar analysis and finally combination of all of those
and with a hint from my husband I was able to use the zoo package with trial 
and error! But it worked and I cleaned out the refunded records.   
•	The most interesting and exciting part of visualization was the mapping. 
Since I have many years of experience of using mapping software such as ArcGIS,
I was really excited to try mapping in R. How ever I found it really 
challenging. I used a shape file of Ohio’s zip codes and every single step took
me a great deal of time. I spend two days on reading different resources and 
finally a combination of online resources and Udacity’s awesome forum mentors 
helped me to make a map.    
•	Another problem that I faced was many times I had to go back to data cleaning
and exclude some records for all analysis or for some of the plots. I think I 
should have study the data in more depth at the beginning to avoid this going 
back and forth.   
•	I spend a good amount of time on thinking about the variables that I have and
their possible relationships as well as what is the best way to present them. 
Trying different plot types, sub setting the data and grouping it and making 
new variables were the strategies that I used.  
•	Using ggplot2 to the extend that I did was really awesome. I feel confident 
that now I can use this package really well. 

### Future Work

> The first thing that comes to mind is probably adding different data from 
reliable sources will take this analysis to the next level and makes it easy 
to accept or reject after election analysis regarding the socio-economic group
of people who donated and voted for Stein, Trump, Clinton, and Sanders. Census
data will be a good resource for this and poverty, race, gender, and level of 
education are the additional data that will help with this analysis. 
